## User's guide {#guide}

This section introduces more features of the `consonance` package.
Code examples require the following packages.

```{r}
library(consonance)
library(checkmate)
library(dplyr)
```

For concreteness, we will use the `anscombe` dataset available with base R.

```{r}
head(anscombe, 2)
```

The dataset holds four pairs of (x, y) coordinates in arbitrary units.
It will be convenient to arrange the data into four separate data frames.

```{r}
datasets <- list(A = data.frame(id = "A", x = anscombe$x1, y = anscombe$y1),
                 B = data.frame(id = "B", x = anscombe$x2, y = anscombe$y2),
                 C = data.frame(id = "C", x = anscombe$x3, y = anscombe$y3),
                 D = data.frame(id = "D", x = anscombe$x4, y = anscombe$y4))
```

What makes these datasets interesting is that they have very similar
(Pearson) correlations, but quite distinct patterns.

```{r anscombe_plots, echo=FALSE, eval=TRUE, fig.width=6.5, fig.height=1.3, dpi=300, out.width="100%"}
oldpar <- par()
par(mfrow=c(1,4))
plot_anscombe(datasets$A)
plot_anscombe(datasets$B)
plot_anscombe(datasets$C)
plot_anscombe(datasets$D)
newpar <- par(oldpar)
```

In this section, we will set up consonance suites for this data and for
regression models based on these four patterns.



### Data consonance

#### Vectors

We saw in the overview how to check that data vectors are numeric.

```{r}
suite_numeric_vec <-
  consonance_suite() +
  consonance_assert("numeric vector", assert_numeric, any.missing=FALSE)
# perform the assessment on dataset "A"
datasets$A$x %>% test_consonance(suite_numeric_vec)
datasets$A$y %>% test_consonance(suite_numeric_vec)
```


#### Data frames & data tables

When data consists of a data frame, data table, or list, test can be carried
out on individual columns / components. This behavior can be triggered via
argument `.var`.

```{r}
suite_numeric_df <-
  consonance_assert("numeric x", assert_numeric, any.missing=FALSE, .var="x") +
  consonance_assert("numeric y", assert_numeric, any.missing=FALSE, .var="y")
```

The argument `any.missing` in each assertion is understood to be passed to
the assertion function, `assert_numeric`. Argument `.var` begins with
a `.` in order to avoid name-clashing with arguments that may be relevant to the
assertion function. It instructs package `consonance` to carry out the assertion
on the designated component.

Applying this suite on one of the `anscombe` datasets should execute quietly.

```{r}
datasets$A %>% test_consonance(suite_numeric_df)
```

To observe a failed test, we can apply the suite on a corrupted data frame.

```{r, error=TRUE}
temp <- datasets$A
temp$x[2] = NA
temp %>% test_consonance(suite_numeric_df)
```

The error messages indicate the corrupt data was detected, as intended.



#### Objects





### Custom assertions

We have seen that consonance assertions can be defined using `assert_numeric`
and `assert_named` from the `checkmate` package. Indeed, `checkmate` provides
dozens of functions that cover many often-used criteria. All of those
functions can be incorporated into a consonance suite. Nonetheless, there may
arise situations where a ready-made assertion function is not available.
In those scenarios it is possible to define a custom assertion.

As an example, let's suppose we want to require that a vector has at least
three distinct values. We can define a custom function.

```{r}
assert_n_unique <- function(x, n) {
  stopifnot(length(unique(x)) >= n)
  x
}
```

Then, we can incorporate this function into a test suite.

```{r}
suite_3_unique <-
  consonance_suite() +
  consonance_assert("at least 3 distinct x", assert_n_unique, n=3, .var="x")
```

Next, we can apply this suite on the `anscombe` datasets.

```{r, error=TRUE}
datasets$A %>% test_consonance(suite_3_unique)
datasets$D %>% test_consonance(suite_3_unique)
```

Dataset "A" passes the consonance assessment, but dataset "D" does not (refer
to the figure). This is not a purely conceptual exercise: if a step in the
analysis relies on `smooth.spline` to fit a model, the x-coordinates will
need to have at least three unique positions for spline nodes. This
procedure detects that smoothing splines will not be feasible for dataset "D".




### Terminology: assert, check, test, expect

There are several related term that appear in the literature on unit testing
and argument checking: assert, check, test, expect. Although they are almost
synonyms in everyday language, they can delineate precise behaviors in code.
This is relevant when adding components into a consonance suite, and
especially when writing custom functions.

The `consonance` package follows the same convention as package `checkmate`
for assert, check, and test functions. Given a function `f(x)`, the three
function types evaluate the data `x` and signal a positive (PASS) or a
negative (FAIL) outcome. The signals are summarized below.

| Function type | outcome: PASS | outcome: FAIL |
| --------------| ---- | ---- |
| assert        | `x` or `invisible(x)` | `stop()` |
| check         | `TRUE` | `"string with error message"` |
| test          | `TRUE` | `FALSE` |

Given this terminology, function `assert_n_unique` is an assertion because it
returns its input or stops execution. An analogous behavior might be written
as a test as follows

```{r}
test_n_unique <- function(x, n) {
  length(unique(x)) >= n
}
```

This alternative function type can also be used within a consonance suite.
However, it should be added via a different constructor.

```{r}
suite_3_unique_alt <-
  consonance_suite() +
  consonance_test("at least 3 distinct x", test_n_unique, n=3, .var="x")
```

Note that this definition involves `consonance_test` rather than
`consonance_assert`. For a check function, the relevant constructor is
`consonance_check`.

All three function types can be used together in a single consonance suite.
However, some unexpected may arise if a function of one type is added
with an inappropriate constructor. It is probably a good idea to pick one
function style and stick with it.

**To do:** discuss functions of type 'expect' and package `testthat`.

*Aside:* natural language also has other verbs like 'verify' and 'validate'
with similar meanings. These do not have precise analog in the package.



### Importance levels

We have seen that `test_consonanace` can generate error messages and halt
execution via R's `stop`. In some situations, it may be appropriate to
signal a potential problem but to nonetheless continue execution.
This can be achieved by setting importance levels using argument `.level`.

As a practical example, let's look at the distribution of x-coordinates
in the `anscombe` dataset.

```{r}
# from the scatter plots, x-coordinates are in range [4, 19]
anscombe.x = unlist(lapply(datasets, function(d) { d$x } ))
anscombe.qs = quantile(anscombe.x, p=c(0, 0.1, 0.9, 1))
anscombe.qs
```

`anscombe.qs` gives the minimum and maximum values for x, and the
10\% and 90\% quantiles as well. If we construct regression models (next
section), it would for relevant to know if a new value for x lies in this
range. Let's make a suite that uses the 10\% and 90\%
quantiles to signal a warning, and the min/max values to escalate to an error.

```{r}
suite_x <-
  consonance_assert("within range", assert_numeric,
                    lower=anscombe.qs[1], upper=anscombe.qs[4],
                    .var="x", .level="error") +
  consonance_assert("within inner range", assert_numeric,
                    lower=anscombe.qs[2], upper=anscombe.qs[3],
                    .var="x", .level="warning")
```

The first assertion uses `.level="error"`, which is the default value and
corresponds to the examples we have seen before. The second assertion uses
`.level="warning"`. Both act on a column `.var="x"` in a data frame. So
let's create small data frames.

```{r, error=TRUE}
# a value of x outside the range should here raise an error
data.frame(x=min(anscombe.x)-1) %>% test_consonance(suite_x)
# a value of x on the outskirts of the distribution should raise a warning
data.frame(x=mean(anscombe.qs[1:2])) %>% test_consonance(suite_x)
```

Note that the last example generates only warnings and allows execution
to continue.



### Model consonance

The package enables attaching a consonance suite to R's 'S3' object, for
example to outputs from `lm`, `glm`, or other modeling frameworks. To
demonstrate this, let's create a regression model for the second `anscombe`
series.

```{r}
B_lm <- lm(y~x, data=datasets$B)
B_lm
```

As the plot at the start of the section shows, a linear regression is a
reasonable first approximation for the series. But the series is actually
better modeled by a parabola; it would be a mistake to use the linear model
outside the x-range used to fit the model. So let's create a suite of
tests to capture this reasoning.

```{r}
suite_lm <- consonance_suite() +
  consonance_assert("x range", assert_numeric, .var="x", .level="warning",
                    lower=min(datasets$B$x), upper=max(datasets$B$x))
```

We can now attach the suite to the model.

```{r}
B_lm_2 <- attach_consonance(B_lm, suite_lm)
```

(This is actually just a long-form syntax, and the same result can be achieved
by setting `B_lm_2$consonance <- suite_lm`.) The power of this step is that
we can now use the model to perform data consonance checks.

```{r}
# prediction on x-coordinates within the model range
in_range <- data.frame(x=c(5, 10))
in_range %>% test_consonance(B_lm_2) %>% predict(object=B_lm_2)
# prediction on x-coordinates outside of the model range
out_range <- data.frame(x=c(20, 30))
out_range %>% test_consonance(B_lm_2) %>% predict(object=B_lm_2)
```

The second dataset generates warnings and produces y-coordinates
that are quite far from what one might guess from a parabolic trend (refer
to the figure).

Note that this example produces warning messages, but it nonetheless displays
a prediction. This is because the assertion in `suite_lm` was defined with
`.level="warning"`. It would also be reasonable to set `.level="error"`
(the default). `test_consonance` would then halt execution before any
prediction is made.




### Logging

By default, `test_consonance` does not generate output when
evaluated on consonant data and outputs messages upon problems. The
level of logging can be adjusted by setting `logging.level` to `"INFO"`,
`"WARN"`, or `"ERROR"`.

Setting the logging level in the suite constructor affects all evaluations.

```{r}
# create a suite that will generate a lot of log messages
suite_verbose <-
  consonance_suite(logging.level="INFO") +
  consonance_assert("character vector", assert_character)
# simple evaluations will generate log messages
test_consonance(c("a", "b"), suite_verbose)
```

Another way to toggle the logging level is within `test_consonance`.

```{r}
# set up a canonical suite
suite_standard <-
  consonance_suite() +
  consonance_assert("character vector", assert_character)
# by default, successful runs will not generate messages
test_consonance(c("a", "b"), suite_standard)
# but we can request more information
test_consonance(c("a", "b"), suite_standard, logging.level="INFO")
# logging goes back to normal without the explicit argument
test_consonance(c("a", "b"), suite_standard)
```

Instead of writing messages to the console, the logger can output into a file
instead. This is achieved with argument `log.file`, either in the suite
constructor or within `test_consonance`. For brevity, an example of the
latter is as follows.

```{r, echo=TRUE, eval=FALSE}
# the verbose suite would normally display messages in the console
test_consonance(c("a", "b"), suite_verbose, log.file="my-log.log")
```

*Note:* If a consonance suite were to fail while using a file-based logger,
messages would appear only in the file and not in the console. Execution
would nonetheless be halted.

