## User's guide {#guide}

This section introduces more features of the `consonance` package.
As in the overview sections, code examples require the following packages.

```{r}
library(consonance)
library(checkmate)
library(magrittr)
```

For concreteness, we will use the `anscombe` dataset available with base R.

```{r}
head(anscombe, 2)
```

The dataset holds four pairs of (x, y) coordinates in arbitrary units.
It will be convenient to arrange the data into four separate data frames.

```{r}
datasets <- list(A = data.frame(id = "A", x = anscombe$x1, y = anscombe$y1),
                 B = data.frame(id = "B", x = anscombe$x2, y = anscombe$y2),
                 C = data.frame(id = "C", x = anscombe$x3, y = anscombe$y3),
                 D = data.frame(id = "D", x = anscombe$x4, y = anscombe$y4))
```

What makes these datasets interesting is that they have very similar
(Pearson) correlations, but quite distinct patterns.

```{r anscombe_plots, echo=FALSE, eval=TRUE, fig.width=6.5, fig.height=1.3, dpi=300, out.width="100%"}
oldpar <- par()
par(mfrow=c(1,4))
plot_anscombe(datasets$A)
plot_anscombe(datasets$B)
plot_anscombe(datasets$C)
plot_anscombe(datasets$D)
newpar <- par(oldpar)
```

In this section, we will set up consonance suites for this data and for
regression models.


### Data consonance

#### Vectors

We saw in the overview how to check that data vectors are numeric.

```{r}
suite_numeric_vec <-
  consonance_suite() +
  consonance_assert("numeric vector", assert_numeric, any.missing=FALSE)
# perform the assessment on dataset A
datasets$A$x %>% test_consonance(suite_numeric_vec)
datasets$A$y %>% test_consonance(suite_numeric_vec)
```


#### Data frames & data tables

When data consists of a data frame, data table, or list, test can be carried
out on individual columns / components. This behavior can be triggered via
argument `.var`.

```{r}
suite_numeric_df <-
  consonance_assert("numeric x", assert_numeric, any.missing=FALSE, .var="x") +
  consonance_assert("numeric y", assert_numeric, any.missing=FALSE, .var="y")
```

The argument `any.missing` in each assertion is understood to be passed to
the assertion function, `assert_numeric`. Argument `.var` begins with
a `.` in order to avoid name-clashing with arguments that may be relevant to the
assertion function. It instructs package `consonance` to carry out the assertion
on the designated component.

Applying this suite on one of the `anscombe` datasets should execute quietly.

```{r}
datasets$A %>% test_consonance(suite_numeric_df)
```

To observe a failed test, we can apply the suite on a corrupted data frame.

```{r, error=TRUE}
temp_df <- datasets$A
temp_df$x[2] = NA
temp_df %>% test_consonance(suite_numeric_df)
```

The error messages indicate the corrupt data was detected, as intended.


#### Lists and other objects

The approach described above to test columns in data frames also applies to
components in arbitrary objects or lists.

```{r, error=TRUE}
# a list object with miscellaneous components, including $x and $y
temp_list_A <- list(x=c(1, 2, 3), y=c(1, 2), comment="x and y numeric")
# testing temp_list_A should execute quietly
temp_list_A %>% test_consonance(suite_numeric_df)
# another list object with a y component that is not numeric
temp_list_B <- list(x=c(1, 2), y=factor(c(1, 2)), comment="y is not numeric")
# testing temp_list_B should generate messages
temp_list_B %>% test_consonance(suite_numeric_df)
```

Although `temp_list_A` and `temp_list_B` are not data frames, the suite
of tests `suite_numeric_df` applies the criteria to components `x` and `y`
just as before. But it is important to disclose
that the `.var` argument in the test constructors only work one-level
deep. It is not possible to specify deep lookups via the constructor, i.e.
it is not possible to request that a test extract component x from
an object, then extract a sub-component x2, and check properties of the
sub-component. To implement such logic, it is necessary to define a
custom assertion.


### Custom assertions

We have seen that consonance assertions can be defined using `assert_numeric`
and `assert_named` from the `checkmate` package. Indeed, `checkmate` provides
dozens of functions that cover many often-used criteria, and all of them
can be incorporated into a consonance suite. Nonetheless, there may
arise situations where a ready-made assertion function is not available.
In those scenarios it is possible to define a custom function.

As an example, let's suppose we want to require that a vector has at least
three distinct values.

```{r}
assert_n_unique <- function(x, n) {
  stopifnot(length(unique(x)) >= n)
  x
}
```

Then, we can incorporate this function into a test suite using `n=3`.

```{r}
suite_3_unique <-
  consonance_suite() +
  consonance_assert("at least 3 distinct x", assert_n_unique, n=3, .var="x")
```

Next, we can apply this suite on the `anscombe` datasets.

```{r, error=TRUE}
datasets$A %>% test_consonance(suite_3_unique)
datasets$D %>% test_consonance(suite_3_unique)
```

Dataset A passes the assessment, but dataset D does not (refer to the
dataset visualizations above). This is not a purely conceptual exercise: if a
step in the analysis relies on `smooth.spline` to fit a model, the
x-coordinates will need to have at least three unique positions for spline
nodes. This procedure detects that smoothing splines will not be feasible for
dataset D.

Another type of custom function is described in the separate vignette on
consonance testing for models.


### Terminology: assert, check, test, etc.

There are several related term that appear in the literature on unit testing
and argument checking: assert, check, test, expect, validate, verify, insist,
ad perhaps others. These terms are almost synonyms in everyday language, and in
the context of code they are sometimes used inconsistently and at other times
delineate precise behaviors.

The `consonance` package uses the word 'test' in a loose sense as well as in
a precise sense. The loose sense is used in the 'consonance testing', in the
label for a component in a suite object (e.g. `suite$tests`), and in the
function `test_consonance`. The word is meant to convey applying criteria on
input data.

The `consonance` package also uses the terms `assert`, `check`, and `test` in
a precise sense, following the convention described in package `checkmate`.
Given a function `f(x)`, the three actions process an object `x` and signal
a positive (PASS) or a negative (FAIL) outcome.

| Function type | outcome: PASS | outcome: FAIL |
| --------------| ---- | ---- |
| assert        | `x` or `invisible(x)` | `stop()` |
| check         | `TRUE` | `"string with error message"` |
| test          | `TRUE` | `FALSE` |

Given this terminology, function `assert_n_unique` is an assertion because it
returns its input or stops execution. An analogous behavior might be written
as a test as follows

```{r}
test_n_unique <- function(x, n) {
  length(unique(x)) >= n
}
```

This alternative function type can also be used within a consonance suite.
However, it should be added via a different constructor.

```{r}
suite_3_unique_alt <-
  consonance_suite() +
  consonance_test("at least 3 distinct x", test_n_unique, n=3, .var="x")
```

Note that this definition involves `consonance_test` rather than
`consonance_assert`. For a check function, the relevant constructor is
`consonance_check`.

All three function types can be used together in a single consonance suite.
However, some puzzling behaviors may arise if a function of one type is added
with an inappropriate constructor. It is probably a good idea to pick one
function style and stick with it.

Apart from assert, check, and test, natural language also has other verbs
with similar meanings. The `testthat` package uses 'expect', the `assertthat`
package uses 'assert that', and the `assertr` package also uses 'verify' and
'insist'. Unfortunately, some of those constructs serve different purposes
than what is provided by the `consonance` package, so they do not have direct
analogs. To mitigate potential confusion, this vignette and other docs try to
avoid using those alternative keywords.


### Importance levels

We have seen that `test_consonanace` can generate error messages and halt
execution via R's `stop`. In some situations, it may be appropriate to
signal a potential problem but to nonetheless continue execution.
This can be achieved by setting importance levels using argument `.level`.

As a practical example, let's look at the distribution of x-coordinates
in the `anscombe` dataset.

```{r}
# from the scatter plots, x-coordinates are in range [4, 19]
anscombe.x = unlist(lapply(datasets, function(d) { d$x } ))
anscombe.qs = quantile(anscombe.x, p=c(0, 0.1, 0.9, 1))
anscombe.qs
```

`anscombe.qs` gives the minimum and maximum values for x, and the
10\% and 90\% quantiles as well. If we construct regression models (next
section), it would for relevant to know if a new value for x lies in this
range. Let's make a suite that uses the 10\% and 90\%
quantiles to signal a warning, and the min/max values to escalate to an error.

```{r}
suite_x <-
  consonance_assert("within range", assert_numeric,
                    lower=anscombe.qs[1], upper=anscombe.qs[4],
                    .var="x", .level="error") +
  consonance_assert("within inner range", assert_numeric,
                    lower=anscombe.qs[2], upper=anscombe.qs[3],
                    .var="x", .level="warning")
```

The first assertion uses `.level="error"`, which is the default value and
corresponds to the examples we have seen before. The second assertion uses
`.level="warning"`. Both act on a column `.var="x"` in a data frame. So
let's create small data frames.

```{r, error=TRUE}
# a value of x outside the range should here raise an error
data.frame(x=min(anscombe.x)-1) %>% test_consonance(suite_x)
# a value of x on the outskirts of the distribution should raise a warning
data.frame(x=mean(anscombe.qs[1:2])) %>% test_consonance(suite_x)
```

Note that the last example generates only warnings and allows execution
to continue. It is also possible to run `test_consonance` in a very strict
mode and stop execution even upon warnings. This can be achieved by setting
`level`.

```{r, error=TRUE}
# a value of x on the outskirts of the distribution should raise a warning.
# by setting level="warning", the test suite will raise an error
data.frame(x=mean(anscombe.qs[1:2])) %>% test_consonance(suite_x, level="warning")
```



### Model consonance

The package enables attaching a consonance suite to any R object, for example
to outputs from `lm`, `glm`, or other modeling frameworks. To demonstrate
this, let's create a regression model for the second `anscombe` series.

```{r}
B_lm <- lm(y~x, data=datasets$B)
B_lm
```

As the plot at the start of the section shows, a linear regression is a
reasonable first approximation for the series, but the data is actually
better modeled by a parabola. It would be a mistake to use the linear model
outside the x-range used to fit the model. So let's create a suite of
tests to capture this reasoning.

```{r}
suite_lm <- consonance_suite() +
  consonance_assert("x range", assert_numeric, .var="x", .level="warning",
                    lower=min(datasets$B$x), upper=max(datasets$B$x))
```

We can now attach the suite to the model.

```{r}
B_lm_2 <- attach_consonance(B_lm, suite_lm)
```

The suite can be previewed through the `$consonance` component of the object.

```{r}
B_lm_2$consonance
```

We can now use the model to perform data consonance checks.

```{r}
# prediction on x-coordinates within the model range
in_range <- data.frame(x=c(5, 10))
in_range %>% test_consonance(B_lm_2) %>% predict(object=B_lm_2)
# prediction on x-coordinates outside of the model range
out_range <- data.frame(x=c(20, 30))
out_range %>% test_consonance(B_lm_2) %>% predict(object=B_lm_2)
```

The second dataset generates warnings and produces y-coordinates
that are quite far from what one might guess from a parabolic trend (refer
to the figure).

Note that this example produces warning messages, but it nonetheless displays
a prediction. This is because the assertion in `suite_lm` was defined with
`.level="warning"`. It would also be reasonable to set `.level="error"`
(the default). `test_consonance` would then halt execution before any
prediction is made.


### Logging

By default, `test_consonance` does not generate output when
evaluated on consonant data and outputs messages upon problems. The
level of logging can be adjusted by setting `logging.level` to `"INFO"`,
`"WARN"`, or `"ERROR"`.

Setting the logging level in the suite constructor affects all evaluations.

```{r}
# create a suite that will generate a lot of log messages
suite_verbose <-
  consonance_suite(logging.level="INFO") +
  consonance_assert("character vector", assert_character)
# simple evaluations will generate log messages
test_consonance(c("a", "b"), suite_verbose)
```

Another way to toggle the logging level is within `test_consonance`.

```{r}
# set up a canonical suite
suite_standard <-
  consonance_suite() +
  consonance_assert("character vector", assert_character)
# by default, successful runs will not generate messages
test_consonance(c("a", "b"), suite_standard)
# but we can request more information
test_consonance(c("a", "b"), suite_standard, logging.level="INFO")
# logging goes back to normal without the explicit argument
test_consonance(c("a", "b"), suite_standard)
```

Instead of writing messages to the console, the logger can output into a file
instead. This is achieved with argument `log.file`, either in the suite
constructor or within `test_consonance`. For brevity, an example of the
latter is as follows.

```{r, echo=TRUE, eval=FALSE}
# the verbose suite would normally display messages in the console
test_consonance(c("a", "b"), suite_verbose, log.file="my-log.log")
```

