## Overview {#overview}

This section gives a preview of the `consonance` package. To start,
let's create a test suite that requires data vectors to be numeric.

```{r}
library(consonance)
library(checkmate)
suite_1 <-
  consonance_suite() +
  consonance_assert("numeric vector", assert_numeric)
```

The first two lines load packages `consonance` and `checkmate`. The latter
provides many fast functions designed for checking arguments.
Next in the code block, the definition of `suite_1` is a
sum of two components. The first is an empty consonance suite created by
`consonace_suite()`. The plus sign and function `consonance_assert()` augment
the suite with a new assertion. Its description, `"numeric vector"`,  is meant
to distinguish this component from any additional ones. The implementation of the
assertion is provided by `assert_numeric` from the `checkmate` package.

To use the `suite_1` object, we invoke the `test_consonance` function. Here are
some examples.

```{r, error=TRUE}
# Testing character vectors results in quiet execution
test_consonance(c(1, 2), suite_1)
test_consonance(c(1, 2, NA), suite_1)
# Testing a non-character vector results in messages
test_consonance(c("a", "b"), suite_1)
```

All examples provide `test_consonance` with a data vector and our test suite.
The first two lines succeed quietly because their data vectors are composed
of numbers. The last line signals that the data is not in agreement, i.e.
not in consonance, with the properties defined in the suite.

Of course, testing that a vector is numeric can be achieved with one
line of code in base R. A more useful suite might combine more conditions.
For example, we can check data type and names.

```{r}
suite_2 <-
  consonance_assert("numeric vec wo NAs", assert_numeric, any.missing=FALSE) +
  consonance_assert("named", assert_named)
```

The first `consonance_assert` definition here includes an additional argument.
This is passed to the assertion functions, i.e. to `assert_numeric` from the
`checkmate` package, and here imposes a constraint that data should not
include any missing values. We can try the suite on some data, this time
using the `dplyr` syntax.

```{r, error=TRUE}
library(dplyr)
# Numeric vectors with names result in quiet execution
c(a=1, a=2) %>% test_consonance(suite_2)
# Numeric vectors without names result in error messages
c(1, 2) %>% test_consonance(suite_2)
# Vectors with missing data result in error messages
c(a=1, b=NA) %>% test_consonance(suite_2)
```

The convenience of the `consonance` workflow becomes even more
apparent when performing long series of tests, testing more complex objects
like data frames or models, logging output, and toggling strictness of
individual requirements.

