## Overview {#overview}

This section gives a preview of the `consonance` package. We will need
the following packages.

```{r}
library(consonance)
library(checkmate)
library(magrittr)
```

In addition to `consonance`, we load `checkmate` as that provides many fast
functions designed for checking arguments, which we can use in a consonance
suite. The `dplyr` package provides the piping operator `%>%`.

For a practical demonstration, let's create a consonance test suite that
requires data vectors to be numeric.

```{r}
suite_1 <- consonance_suite() +
  consonance_assert("numeric vector", assert_numeric)
```

This definition is a sum of two components. The first is an empty
consonance suite created by `consonace_suite()`. The plus sign and function
`consonance_assert()` augment the suite with a new assertion. Its description,
`"numeric vector"`,  is meant to distinguish this component from any
additional ones. The implementation of the assertion is provided by
`assert_numeric` from the `checkmate` package.

To use the object, we invoke the `test_consonance` function. Here are some
examples.

```{r, error=TRUE}
# Testing numeric vectors results in quiet execution
test_consonance(c(1, 2), suite_1)
test_consonance(c(1, 2, NA), suite_1)
# Testing a non-numeric vector results in messages
test_consonance(c("a", "b"), suite_1)
```

All examples provide `test_consonance` with a data vector and our test suite.
The first two lines succeed quietly because their data vectors are composed
of numbers. The last line signals that the data is not in agreement, i.e.
not in consonance, with the criteria in the suite.

Of course, testing that a vector is numeric can be achieved with one
line of code in base R. The consonance suite framework becomes more useful
when we combine more conditions. For example, we can check data type and names.

```{r}
suite_2 <-
  consonance_assert("numeric vec wo NAs", assert_numeric, any.missing=FALSE) +
  consonance_assert("named", assert_named)
```

The first `consonance_assert` definition here includes an additional argument.
This is passed to the assertion function, i.e. to `assert_numeric` from the
`checkmate` package. Here, it imposes a constraint that data should not
include any missing values.

We can try the suite on some data, this time using the `dplyr` syntax.

```{r, error=TRUE}
# Numeric vectors with names result in quiet execution
c(a=1, a=2) %>% test_consonance(suite_2)
# Numeric vectors without names result in error messages
c(1, 2) %>% test_consonance(suite_2)
# Vectors with missing data result in error messages
c(a=1, b=NA) %>% test_consonance(suite_2)
```

The convenience of the `consonance` workflow becomes even more
apparent when performing long series of tests, testing more complex objects
like data frames or models, logging output, and toggling strictness of
individual requirements.

