## Use cases {#usecases}

As mentioned in the introduction, the `consonance` package uses capabilities
provided by the base R environment and existing packages to make it easier
to define, maintain, and deploy quality controls for data. While the user's
guide introduced these features one at a time, here we can look at practical
use cases. These scenarios are relevant to software development, data
analysis by one analyst, and deploying models for others to use.


### Argument checking

Function definitions often implement (or should implement) some checks on
their arguments. This sometimes leads to duplicated code when a family of
functions may need to test arguments for the same conditions. At the same
time, some functions may need to operate both in 'safe-mode' with tests
activated and in a 'fast-mode' with tests turned off. Both these issues
can be address with consonance testing: by re-using one suite of tests
in more than one custom function, and by providing a switch to execute or to
skip testing.

```{r, echo=TRUE, eval=FALSE}
# pseudo-code
suite_arguments <- ...
custom_function_1 <- function(x, skip.qc=FALSE) {
  validate(x, suite_arguments, skip=skip.qc)
  # work for custom function 1
}
custom_function_2 <- function(x, skip.qc=FALSE) {
  validate(x, suite_arguments, skip=skip.qc)
  # work for custom function 2
}
```


### Data quality control

Consonance testing can act as a tool to help investigate state during
the course of a pipeline and to log progress.

```{r, echo=TRUE, eval=FALSE}
# pseudo-code for a hypothetical pipeline - standard use
large_data %>%
  validate(suite_pipeline) %>%
  work_on_data_1 %>%
  validate(suite_pipeline) %>%
  work_on_data_2
# pseudo-code for a pipeline - manual toggling for speed or debugging
large_data %>%
  validate(suite_pipeline, skip=TRUE) %>%
  work_on_data_1 %>%
  validate(suite_pipeline, logging.level="INFO") %>%
  work_on_data_2
```

Note that stages in the pipeline with consonance testing can be easily
skipped using the `skip` argument, made more-or-less verbose using logging,
or entirely added/removed with no side effects on the remaining pipeline
workflow.


### Model and data applicability

Attaching a consonance suite to a model allows a data scientist help
end-users apply the model within its intended range of applicability.
Consonance suites can be included with any model, saved into an `Rda` file
using `save`, and applied in new R sessions.

```{r, echo=TRUE, eval=FALSE}
# pseudo-code for modeler
model <- train_a_model_on_data(large_data)
model$consonance <- suite_for_model
save(model, file="model.Rda")
# pseudo-code for end-user
load("model.Rda")
new_data %>% validate(model) %>% predict(object=model)
```

This is particularly useful to raise warnings for data that is technically
well-formatted, but may not adhere to the assumptions made during model
training.

A related application is to include a test suite with a dataset that is
distributed for re-analysis. Before an existing dataset is integrated or
jointly analyzed with new data, a test suite associated with the original
dataset can signal whether or not the new data has compatible properties.
Unlike for models, however, this use case requires distribution of a test
suite or data-merging toolkit that is separate from the data.

