---
title: Consonance testing in batch
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Consonance testing in batch}
  %\usepackage[UTF-8]{inputenc}
---

```{r, child=file.path("sections", "styles.Rmd")}
```

```{r, echo=FALSE}
suppressMessages(library(Rcssplot))
library(checkmate)
```


## Introduction

In some situations, it is relevant to perform validations multiple times.

For example, in interactive sessions in which objects can become
over-written - by design or by mistake -  we might want to check them
in the console. Another use case arises in scripts that produce several
objects with overlapping requirements.


## Individual validations

In order to prep for the next section on validation in batch, let's
first review consonance validation on individual objects. Let's set up
a consonance suite for data frames.

```{r}
library(consonance)
suite_numeric <-
  consonance_assert("x numeric", assert_numeric, .var="x", any.missing=FALSE) +
  consonance_assert("y numeric", assert_numeric, .var="y", any.missing=FALSE)
```

For data objects, we will use two small data frames.

```{r}
d1 <- data.frame(x=c(3, 4), y=c(7, 0), z=0)
d1
d2 <- data.frame(x=c(3, NA), y=c("7", "0"), stringsAsFactors=FALSE)
d2
```

The direct way of using consonance is to validate each object
against the suite, one at a time.

```{r, error=TRUE}
# should execute quietly
validate(d1, suite_numeric)
# should signal errors
validate(d2, suite_numeric)
```


## Validations in batch

To perform validations in batch, we first specify pairings of objects and
test-suites.

```{r}
batch <- data.frame(object=c("d1", "d2"), suite="suite_numeric",
                    stringsAsFactors=FALSE)
batch
```

Then, we can validate using this group object

```{r, error=TRUE}
validate(batch)
```

The batch can be of any size, and becomes increasingly useful when working
in environments with a large number of objects. To extend the above example,
we can prepare several other data frames.

```{r}
# additional objects with numeric data
d3 <- data.frame(x=1:3, y=1:3)
d4 <- data.frame(x=1:4, y=1:4)
d5 <- data.frame(x=1:5, y=1:5)
batchB <- data.frame(object=paste0("d", c(1,3,4,5)),
                     suite="suite_numeric",
                     stringsAsFactors=FALSE)
# should execute quietly to signal consonant data
validate(batchB)
````

Quiet execution signals that all data frames pass all the criteria.


## Appendix

```{r}
sessionInfo()
```

