---
title: Consonance testing
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Consonance testing}
  %\usepackage[UTF-8]{inputenc}
---

```{r, child=file.path("sections", "styles.Rmd")}
```

```{r, echo=FALSE}
suppressMessages(library(Rcssplot))
source("plot_consonance.R")
```


## Technical synopsis

Consonance testing refers to the automated checking of compatibility of
data with a processing pipeline.

Several R packages already provide frameworks for unit testing,
argument checking, and assertions. Package `consonance` provides an
intermediate layer between practical data analysis and those core frameworks.
Its unique features include:

 - Tests are grouped into suites that can be re-used many times.
 - Test definitions can use established frameworks (e.g. `checkmate`) as
 well as custom functions.
 - Test suites can be attached to model objects (e.g. `lm`, `glm`).
 - Test suites are coupled to a logger. Output messages can be directed to
  the console or to a log file.
 - Testing can be toggled on and off with a single argument.


```{r, child=file.path("sections", c("introduction.Rmd", "overview.Rmd"))}
```

```{r, child=file.path("sections", c("guide.Rmd", "usecases.Rmd"))}
```



## Appendix {#appendix}

```{r}
sessionInfo()
```

